Learning how to act: making good decisions with machine learning

### 2017-09-10 thesis/thesis_map.tex
fixed grammer issues

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty. My other contributions are primarily connective rather than technical. I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias. I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data, and covariate shift. I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.

@@

Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks. Therefor, in this section, we briefly review the key properties of Bayesian networks. A more thorough introduction (including proofs) can be found in \citep{Koller2009}. Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way which highlights conditional independencies between them.

@@

The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$. D-separation is an algorithm that extends the local Markov property to find these additional independencies. It provides us with a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.

The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ gives us no additional information about $X$. From a graphical perspective you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network. Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$. In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise. In figure (d) the path is unblocked if we condition on $Z$ and blocked otherwise.

The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure. The somewhat counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}. As an example, consider a scholarship available to female or disadvantaged students. Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship. There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent. However, if we know a student is receiving a scholarship then learning that they are male  increases the probability that they are disadvantaged.
\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y| Z$}
\label{fig:dseptripple1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$X$};
\node[main node, below =of 1](2){$Z$};
\node[main node, below =of 2](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    (2) edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|Z$}
\label{fig:dseptripple2}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Y$};
\node[main node, below =of 1](2){$Z$};
\node[main node, below =of 2](3){$X$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    (2) edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|Z$}
\label{fig:dseptripple3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Z$};
\node[main node, below left =of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
        edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|\{\}$}
\label{fig:dseptripple4}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Z$};
\node[main node, above left =of 1](2){$X$};
\node[main node, above right =of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge node {} (1)
    (3) edge node {} (1);	
\end{tikzpicture}
\end{subfigure}
\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
\label{fig:dseptriples}
\end{figure}

@@

If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$. However, it is possible to construct distributions that do not have a perfect map, that is they contain conditional independencies that cannot be represented by d-separation. A particular case in which this occurs is when there are deterministic relationships between variables. If we have a Bayesian network $G$ in which we specify that some nodes are deterministic we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent. This does not conflict with theorem \ref{thrm:dsep_independence} as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables. This subtlety led to confusion in assessing what independencies hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connectedness to assert lack of independence. D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.

@@

The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$. Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$. However setting $X$ tells us nothing about its parents and therefor affects $Y$ only via direct paths out of $X$. Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$. If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths. This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).

@@

\end{tikzpicture} 
\end{subfigure}
\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$. In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$ which also influences $Y$. If we condition on $W$, we block this indirect path.}
\label{fig:rule2}
\end{figure}

@@

The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$. These are the only paths by which intervening in $\boldsymbol{X}$ can effect $\boldsymbol{Y}$.

@@

The question of non-parametric identifiability is solved. The do calculus is complete \citep{Shpitser2006b,Huang2006}. A problem is identifiable if and only if the interventional distribution of interest can be transformed into term containing only observational quantities via repeated application of the do-calculus. There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:

\begin{enumerate}
\item determine if the query can be translated into an expression involving only distributions over observed variables. In other words, determine if the query is identifiable given the assumptions encoded by the network, and
\item if it is identifiable, return the required expression. 
\end{enumerate}

Figure \ref{fig:identify} shows some examples of identifiable and non identifiable queries. I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.

@@

Many interesting questions relating to identifiability remain open. What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable? What if we assume various restrictions on the functional form of the relationships between the variables? Some queries which are not identifiable non-parametrically can be identified by additional assumptions such as linearity. A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.

@@

Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function. There are a variety of ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions. For a binary treatment variable $X$, the average causal effect, ACE \footnote{also referred to as the average treatment effect (ATE)} is defined as:

@@

The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}. Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution. However we can personalise the average causal effect by stating it with respect to some observed context. I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE), however that term is also used for the sample rather than population effect.}

@@

In some cases we may be interested in the average causal effect for some sub-group of the population. A particularly common example of this is the average treatment effect of the treatment of the treated (ATT). This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.

@@

The backdoor criterion derives from rule 2 of the do-calculus. Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set which satisfies the backdoor criterion. There may be more than one valid adjustment set, see figure \ref{fig:adjust_multiple}. The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}. Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}. Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for. For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether birth weight should be adjusted for to estimate the causal effect of smoking on neonatal mortality.

\begin{figure}
\centering
\begin{subfigure}[t]{0.48\textwidth}
\centering
\caption{Multiple valid adjustment sets.}
\label{fig:adjust_multiple}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](2){$X$};
\node[ob, right=of 2](3){$Y$};
\node[ob,above=of 3](4){$Z_2$};
\node[ob,above=of 2](5){$Z_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (5) edge (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.48\textwidth}
\centering
\caption{Don't condition on all observable variables}
\label{fig:m-graph}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1);
\end{tikzpicture} 
\end{subfigure}
\caption{Identifying an optimal adjustment set is not always intuitively obvious. There may be multiple valid adjustment sets. In sub-figure (a) $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$. Adding additional variables (even pre-treatment variables) to a valid adjustment set can result in an invalid set as in sub-figure (b). In this case the empty set is a valid adjustment set for the causal effect of $X$ on $Y$, however adding $Z$ would open a backdoor path $X-U_1-Z-U_2-Y$}
\label{fig:adjustment}
\end{figure}

@@

Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata. However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates. There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability, see \citet{Imbens2004} for a comprehensive review. The key approaches are based on matching on covariates, propensity score methods and regression. We now examine these approaches from a machine learning perceptive.

@@

We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, we only have access to a sample. We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.

@@

\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem. If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$ but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$. \footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift. Take the case where we have a binary intervention $x \in \set{0,1}$. Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$. More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.

@@

The average causal effect can then be estimated by,

@@

The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}. Doubly robust methods are asymptotically unbiased as long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.

The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$. \citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$. This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$. Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}. In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013} the propensity score may be known. However in general, it must be estimated from data. Frequently this is done with a simple parametric model such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}. \citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.

@@

This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression. There are many variants of matching estimators utilising different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}. Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}. An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.

The practical performance of the estimation approaches discussed in this section will depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions. A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.

@@

The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}. Unfortunately there are a relatively small number of examples where comparable observational and experimental data are available. The results are mixed with later studies finding generally better alignment of results but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data-sets.

>>>>

\paragraph{Contributions} The goal of this thesis is to connect and unify the key approaches to solving causal problems from both the observational and interventional viewpoints. My major contribution is a framework that unifies the causal graphical model approach for inference in observational settings with the sequential experimental approach encapsulated by multi-armed bandits. This framework allows us to represent knowledge of how variables are related to one-another in a very natural way and induces an interesting and novel form of structure between the different actions modelled in the bandit problem. I develop a new algorithm that can exploit this structure as a first step towards a unified approach to decision making under uncertainty. 
I make a number of additional connective contributions that are not encompassed by my work on causal bandit problems. I demonstrate the role of a formal causal framework within Bayesian approaches to inference and show how assigning a prior based on human causal intuition without considering the causal structure of the problem can introduce bias. I highlight the connections between approaches to off-policy evaluation in bandit problems, causal effect estimation from observational data, and covariate shift. I clarify the implicit causal structure underlying various bandit settings and the counterfactual nature of \emph{regret} - the measure by which bandit algorithms are assessed.

@@

Many causal algorithms are based on leveraging the independence properties encoded in Bayesian networks. Therefore, in this section, we briefly review the key properties of Bayesian networks. A more thorough introduction (including proofs) can be found in \citep{Koller2009}. Recall that a Bayesian network is a way of representing the joint distribution over its variables in a way that highlights conditional independencies between them.

@@

The set of conditional independence relations given by the local Markov condition can enforce additional independencies that also hold in all distributions that are compatible with $G$. D-separation is an algorithm that extends the local Markov property to find these additional independencies. It provides a simple way of reading from a network if a given conditional independence statement is true in all distributions compatible with that network.

The statement that $X$ is conditionally independent of $Y$ given $Z$ implies that if we know $Z$, learning the value of $Y$ provides no additional information about $X$. From a graphical perspective, you can think of this as $Z$ blocks the flow of information from $X$ to $Y$ in the network. Figure \ref{fig:dseptriples} shows all possible network paths from a variable $X$ to $Y$ via $Z$. In figures (a) to (c) the path is blocked if we condition on $Z$ and unblocked otherwise. In figure (d), the path is unblocked if we condition on $Z$ and blocked otherwise.

The structure in figure \ref{fig:dseptripple4} is referred to as a collider or v-structure. The counter-intuitive result that conditioning on $Z$ introduces dependence between $X$ and $Y$ is called the \textit{explaining away phenomena}. As an example, consider a scholarship available to female or disadvantaged students. Let $X$ be gender, $Y$ be family background and $Z$ receipt of the scholarship. There are roughly equal numbers of boys and girls in both poor and wealthy families so $X$ and $Y$ are independent. However, if we know a student is receiving a scholarship, then learning that they are male  increases the probability that they are disadvantaged.
\begin{figure}[h]
\centering
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y| Z$}
\label{fig:dseptripple1}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$X$};
\node[main node, below =of 1](2){$Z$};
\node[main node, below =of 2](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    (2) edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|Z$}
\label{fig:dseptripple2}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Y$};
\node[main node, below =of 1](2){$Z$};
\node[main node, below =of 2](3){$X$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
    (2) edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|Z$}
\label{fig:dseptripple3}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Z$};
\node[main node, below left =of 1](2){$X$};
\node[main node, below right=of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (1) edge node {} (2)
        edge node {} (3);	
\end{tikzpicture}
\end{subfigure}
\begin{subfigure}[t]{0.2\textwidth}
\centering
\caption{$X \ci Y|\{\}$}
\label{fig:dseptripple4}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=.5cm,
  thick,main node/.style={observed}]
\node[main node](1){$Z$};
\node[main node, above left =of 1](2){$X$};
\node[main node, above right =of 1](3){$Y$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge node {} (1)
    (3) edge node {} (1);	
\end{tikzpicture}
\end{subfigure}
\caption{All possible two edge paths from $X$ to $Y$ via $Z$}
\label{fig:dseptriples}
\end{figure}

@@

If $\indep{G} = \indep{P}$ then $G$ is called a perfect map for $P$. However, it is possible to construct distributions that do not have a perfect map; that is, they contain conditional independencies that cannot be represented by d-separation. The presence of deterministic relationships between variables is one instance where this occurs. If there is a Bayesian network $G$ with some deterministic nodes, then we cannot conclude that if $X$ and $Y$ are d-connected then there exists a distribution $P'$ \emph{consistent} with $G$ in which they are dependent. This does not conflict with theorem \ref{thrm:dsep_independence}, as \emph{consistent} in this setting requires that $P'$ both factorises over $G$ and satisfies the specified the deterministic relations between variables. This subtlety led to confusion in the independencies that hold between counterfactuals via twin networks \citep{Pearl2000,Richardson2013} and demonstrates the caution required in using d-connectedness to assert lack of independence. D-separation can be extended to compute the additional independencies implied by a graph in which certain nodes are known to be deterministic \citep{Geiger1990}.

@@

The intuition behind this is that interventional distributions differ from observational ones due to the presence of indirect paths between $X$ and $Y$. Observing a variable $X$ provides information about $Y$ both directly and indirectly, by changing our belief about the distribution of the parents of $X$. However, setting $X$ tells us nothing about its parents and therefore affects $Y$ only via direct paths out of $X$. Removing edges \emph{leaving} $X$ removes all the direct paths out of $X$. If $X$ is then independent of $Y$ (conditional on $W$), that indicates there are no indirect paths. This implies conditioning on $X$ is equivalent to setting $X$ (given $W$).

@@

\end{tikzpicture} 
\end{subfigure}
\caption{An example of rule 2 with a single intervention $(Y \ci X|W)$ in  $G_{\underline{\boldsymbol{X}}} \implies \P{Y|do(X),W} = \P{Y|X,W}$. In this example, observing $X$ provides information about $Y$ both directly and indirectly, because knowing $X$ tells us something about $W$, which also influences $Y$. If we condition on $W$, we block this indirect path.}
\label{fig:rule2}
\end{figure}

@@

The statement that $\boldsymbol{Y} \ci \boldsymbol{\hat{X}}$ (without conditioning on $\boldsymbol{X}$) implies that there is no unblocked path from $\boldsymbol{X}$ to $\boldsymbol{Y}$ in $G$ which \emph{includes} an arrow leaving $\boldsymbol{X}$. These are the only paths by which intervening in $\boldsymbol{X}$ can affect $\boldsymbol{Y}$.

@@

The question of non-parametric identifiability is solved. The do calculus is complete \citep{Shpitser2006b,Huang2006}. A problem is identifiable if and only if the interventional distribution of interest can be transformed into terms containing only observational quantities via repeated application of the do-calculus. There is a polynomial time algorithm \citep{Shpitser2006a} based on these properties that, for a given network and interventional (do-type) query, can:

\begin{enumerate}
\item determine if the query can be translated into an expression involving only distributions over observed variables. In other words, it can determine if the query is identifiable given the assumptions encoded by the network, and
\item if it is identifiable, return the required expression. 
\end{enumerate}

Figure \ref{fig:identify} shows some examples of identifiable and non-identifiable queries. I have created a javascript implementation of the identifiability algorithm \citep{Shpitser2006a} on which you can test your own queries http://finnhacks42.github.io.

@@

Many interesting questions relating to identifiability remain open. What is the minimal (by some metric) additional information that would be required to make a non-identifiable query identifiable? What if we assume various restrictions on the functional form of the relationships between the variables? Some queries that are not identifiable non-parametrically can be identified by additional assumptions, such as linearity. A complete algorithm for the problem of linear identifiability is yet to be found, despite a rich body of work \citep{Chen2016,tian2009parameter,drton2011global}.

@@

Frequently however, studies wish to define and estimate a causal effect without reference to a specific utility function. There are several ways of defining causal effects that can be viewed as different ways of summarising the family of interventional distributions. For a binary treatment variable $X$, the average causal effect, ACE \footnote{Also referred to as the average treatment effect (ATE)} is defined as:

@@

The average causal effect is often introduced as the average over individual causal effects as discussed in section \ref{sec:counterfactuals}. Individual causal effects are deterministic and cannot be expressed as properties of the interventional distribution. However, we can personalise the average causal effect by stating it with respect to some observed context. I will refer to this as the personalised causal effect (PCE).\footnote{This quantity is sometimes called the conditional average treatment effect (CATE). However, that term is also used for the sample rather than population effect.}

@@

In some cases, the average causal effect for some sub-group of the population is of prime interest. A particularly common example of this is the average treatment effect of the treatment of the treated (ATT). This would be the key quantity of interest if we had to decide whether or not to continue providing a program or treatment for which we could not control the treatment assignment process.

@@

The backdoor criterion derives from rule 2 of the do-calculus. Selecting which covariates should be adjusted for to estimate a causal effect reduces to identifying a set that satisfies the backdoor criterion. There may be more than one valid adjustment set, (figure \ref{fig:adjust_multiple}). The seemingly simple problem of determining if a variable should be adjusted for when estimating causal effects has been the subject of substantial debate and controversy \citep{pearl2009myth}. Adjusting for the wrong variables (even pre-treatment variables) can introduce or magnify bias, see figure \ref{fig:m-graph}. Causal graphs and the back door criterion provide a clear mechanism for deciding which variables should be adjusted for. For a practical example, see the discussion in \citet{schisterman2009overadjustment} on whether estimates of the causal effect of smoking on neonatal mortality should adjust for birth weight.

\begin{figure}
\centering
\begin{subfigure}[t]{0.48\textwidth}
\centering
\caption{Multiple valid adjustment sets.}
\label{fig:adjust_multiple}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](2){$X$};
\node[ob, right=of 2](3){$Y$};
\node[ob,above=of 3](4){$Z_2$};
\node[ob,above=of 2](5){$Z_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (5) edge (2) edge (4)
    (4) edge (3);
\end{tikzpicture} 
\end{subfigure}
\begin{subfigure}[t]{0.48\textwidth}
\centering
\caption{Don't condition on all observable variables}
\label{fig:m-graph}
\begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=1cm,
  thick,lt/.style={latent},ob/.style={observed}]
\node[ob](1){$Z$};
\node[ob, below left=of 1](2){$X$};
\node[ob, below right=of 1](3){$Y$};
\node[lt,above=of 3](4){$U_3$};
\node[lt,above=of 2](5){$U_1$};
 \path[every node/.style={font=\sffamily\small}]
    (2) edge (3)
    (4) edge (1) edge (3)
    (5) edge (2) edge (1);
\end{tikzpicture} 
\end{subfigure}
\caption{Identifying an optimal adjustment set is not always intuitively obvious. There may be multiple valid adjustment sets. In sub-figure (a) $Z_1$ or $Z_2$ or $\set{Z_1,Z_2}$ all block the backdoor path from $X$ to $Y$. Adding additional variables (even pre-treatment variables) to a valid adjustment set can result in an invalid set as in sub-figure (b). In this case the empty set is a valid adjustment set for the causal effect of $X$ on $Y$. However adding $Z$ would open a backdoor path $X-U_1-Z-U_2-Y$}
\label{fig:adjustment}
\end{figure}

@@

Assuming $x$ and $\vb{z}$ are discrete, equation \ref{eqn:expected_post_intervention}, and thus the ACE, can be estimated by selecting the data for which $X=x$, stratifying by $\vb{Z}$, then computing the mean outcome within each stratum and finally weighting the results by the number of samples in each strata. However this approach is not workable for most real world problems with finite samples as the number of strata grows exponentially with the dimension of $\vb{Z}$ and it cannot handle continuous covariates. There is a substantial body of work within in the statistics and econometrics literature on estimating average causal effects assuming conditional ignorability (see \citet{Imbens2004} for a comprehensive review). The key approaches are based on matching on covariates, propensity score methods and regression. We now examine these approaches from a machine learning perceptive.

@@

We cannot directly compute the generalisation error as $\P{\vb{x},y}$ is unknown, as we only have access to a sample. We could search over $\mathcal{H}$ and select a hypothesis $h^*(\vb{x})$ that minimises some loss function on the sample data.

@@

\citet{Johansson2016_counterfactual_rep} identified that this is equivalent to the covariate shift problem. If we let $\vb{v} = (\vb{x},\vb{z})$ then we have training data sampled from $\Pns{train}{\vb{v}}\P{y|\vb{v}}$ where $\Pns{train}{\vb{v}}=\P{\vb{z}}\P{\vb{x}|\vb{z}}$, but at test time the data will be sampled from $\Pns{test}{\vb{v}}\P{y|\vb{v}}$, where $\Pns{test}{\vb{v}} = \delta(\vb{x}-\vb{x'})\P{\vb{z}}$. \footnote{It is not obvious that the question of estimating causal effects under ignorability entirely reduces to  covariate shift. Take the case where we have a binary intervention $x \in \set{0,1}$. Suppose we learn $h(1,\vb{z}) = \E{Y|x=1,\vb{z}}+g(\vb{z})$ and $h(0,\vb{z}) = \E{Y|x=0,\vb{z}}+g(\vb{z})$, then the estimated average causal effect equals the true average causal effect for any function $g$, $\E{h(1,\vb{z})-h(0,\vb{z})} = \E{Y|x=1,\vb{z}} - \E{Y|x=0,\vb{z}}$. More generally, if the goal is to select an optimal action $x^*$ from a continuous space of possible interventions we need algorithms capable of leveraging any structure in the relationship between $x$ and $y$ as well as a means of focusing the loss on regions of the sample likely to affect $x^*$.} With this connection to covariate shift in mind, let us return to regression, matching and propensity scores.

@@

The average causal effect can then be estimated by:

@@

The estimator in equation \ref{eqn:importance_weighted_erm} is an example of a doubly robust estimator \citep{scharfstein1999adjusting,}. Doubly robust methods are asymptotically unbiased so long as either the regression model $h$ or propensity score $e$ are correctly specified \citep{robins1995semiparametric}.

The propensity score can be used to estimate the average causal effect without specifying a regression model for $\E{Y|X,\vb{Z}}$. \citet{Rosenbaum1983} demonstrated that if the ignorability assumption is satisfied by conditioning on $\vb{Z}$, then it is also satisfied by conditioning on $e(\vb{z})$. This allows for estimators based on stratifying, matching or regression on the propensity score rather than the covariates $\vb{Z}$. Inverse propensity weighting can also be combined with empirical estimation of $\E{Y|X,Z}$ yielding the simple, albeit inefficient, estimator in equation \ref{eqn:simple_inverse_propensity} \citep{Imbens2004}. In some settings, such as stratified randomised trials \citep{imbens2015causal} or learning from logged bandit feedback \citep{Bottou2013}, the propensity score may be known. However, in general, it must be estimated from data. Frequently this is undertaken with a simple parametric model, such as logistic regression, but a wide range of standard machine learning algorithms including bagging and boosting, random forests and neural networks can also be applied   \citep{Austin2011}. \citet{Lunceford2004} review the theoretical properties of key propensity score based estimators, including stratification and inverse propensity weighting.

@@

This estimator is equivalent to equation \ref{eqn:cef_regression_missing} with k nearest neighbour regression. There are many variants of matching estimators using different distance metrics, matching with or without replacement (and in the latter case, greedy or optimal matching) and with or without discarding matches beyond some threshold \citep{Cochran1973,rosenbaum1985constructing}. Although intuitive, matching estimators in general have poor large sample properties \citep{Abadie2006}. An exception is where the goal is to estimate the average treatment effect of treatment on the treated in settings where there is a large set of control instances (compared to treatment instances) \citep{Imbens2004}.

The practical performance of the estimation approaches discussed in this section depend on the sample size, dimensionality of the covariates, the complexity of the treatment assignment mechanism and output function, and the degree of prior knowledge available about these functions. A key difference between standard machine learning problems and causal effect estimation is that when estimating causal effects we cannot directly apply cross-validation or a hold-out set for model selection because we lack samples from the counterfactual.

@@

The performance of methods for causal effect estimation can be tested on simulated data \citep{Frolich2001,zhao2004using,Hill2011,Dorie2016} or by comparing estimates from observational studies with the results from corresponding experiments \citep{lalonde1986evaluating,Fraker1987,heckman1997matching,heckman1998characterizing,dehejia1999causal,Smith2001,Anglemyer2014}. Unfortunately, there are a relatively small number of examples where comparable observational and experimental data are available. The results are mixed with later studies finding generally better alignment of results, but it is hard to ascertain if this is due to improved methodological approaches or over-fitting to the available data.

\section{Bayesian reasoning and causality}
\label{sec:bayesian_causality}
The Bayesian approach to modelling encapsulates a very general approach to combining prior knowledge and assumptions with data to draw inferences. You can specify a complex model that captures your assumptions about the underlying mechanisms of your system, describe uncertainty about the parameters (and priors over their values), and then update your beliefs about the values of these parameters based on the observed data. If you believe the model to be a mechanistic representation of the way the world works then it can be used to infer the consequences of interventions.
